Conservation and losses of avian non-coding RNA loci

### 2014-05-30 paper/manuscript.tex
Added Maria\47s edits.

\begin{abstract}
Here we present the results of a large-scale bioinformatic annotation
of non-coding RNA loci in 48 avian genomes. Our approach uses
probabilistic models of hand-curated families from the Rfam database
to infer conserved RNA families within each avian genome. We
supplement these annotations with predictions from the tRNA annotation
tool, tRNAscan-SE and microRNAs from miRBase.  We show a number of
lncRNA-associated loci are conserved between birds and mammals,
including several intriguing cases where the reported mammalian lncRNA
function is not conserved in birds.  We also demonstrate extensive
conservation of classical ncRNAs (e.g., tRNAs) and more recently
discovered ncRNAs (e.g., snoRNAs and miRNAs) in birds. Furthermore, we
describe numerous ``losses'' of several RNA families, and attribute
these to genuine loss, divergence or missing data.  In particular, we
show that many of these losses are due to the challenges associated
with assembling Avian microchromosomes. These combined results illustrate the utility of
applying homology-based methods for annotating novel vertebrate
genomes.
\end{abstract}

@@

New sequencing technologies have dramatically expanded the rate at
which ncRNAs are discovered and their functions are determined
\cite{cech2014noncoding}. However, in order to determine the full
range of ncRNAs across multiple species we require multiple RNA
fractions (e.g. long and short), in multiple species, in multiple
developmental stages and tissues types. The costs of this approach are
still prohibitive in terms of researcher-time and
finances. Consequently, in this study we concentrate on bioinformatic
approaches, primarily homology based methods (i.e. covariance models
(CMs)). However, we do validate the majority of these predictions
using RNA-seq. The bioinformatic approaches we use remain state of the art for ncRNA bioinformatic
analyses \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009} and have well
established sensitivity and specificity rates \cite{Freyhult:2007}.
For example, the CM based approach for annotating ncRNAs in genomes requires
reliable alignments and consensus secondary structures of
representative sequences of RNA families, many of which can be found
at Rfam
\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}. These
are used to train probabilistic models that score the likelihood that
a database sequence is generated by the same evolutionary processes as
the training sequences based upon both sequence and structural
information \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.  The
tRNAscan-SE software package uses CMs to accurately predict transfer
RNAs \cite{Lowe:1997,Chan:2009}.

Independent benchmarks of bioinformatic annotation tools have shown
that the CM approaches out-perform alternative methods
\cite{Freyhult:2007}, although its sensitivity can be limited for
rapidly evolving families such as vault RNAs or telomerase RNA
\cite{Menzel:09a}.

@@

The NBR2 lncRNA and BRCA1 gene share a bidirectional promotor
\cite{Xu:1997}. Both are expressed in a broad range of
tissues. Extensive research on BRCA1 has shown that it is involved in
DNA repair \cite{Moynahan:1999}. The function of NBR2 remains unknown,
yet its conservation across the vertebrates certainly implies a
function (See Figure~\ref{fig:1}). Possibly at the DNA level, however,
function at the RNA level cannot be ruled out at this stage.

@@

In order to determine to what extent the absence of these RNAs from
the infernal-based annotation is caused by sequence divergence beyond
the thresholds of the Rfam CMs, we complemented our analysis by
dedicated searches for a few of these RNA groups. Our ability to find
additional homologs for several RNA families that fill gaps in the
abundance matrices (Figure~1) strongly suggests that conspicuous
absences, in particular of LUCA and LECA RNAs, are caused by
incomplete data in the current assemblies and sequence divergence
rather then genuine losses.

@@

In this work we have provided a comprehensive annotation of non-coding
RNAs in genome sequences using homology-based methods. The
homology-based tools have distinct advantages over experimental-based
approaches as not all RNAs are expressed in any particular tissue-type
or developmental-stage, in fact some RNAs have extremely specific
expression profiles, e.g. the lsy-6 microRNA \cite{Johnston:2003}.  We
have identified previously unrecognised conservation of ncRNAs in
avian genomes and some surprising ``losses'' of otherwise well
conserved ncRNAs. We have shown that most of these losses are due to
difficulties assembling avian microchromosomes rather than \emph{bona
  fide} gene loss. A large fraction of our annotations have been
confirmed using RNA-seq data, which also showed a 7-fold enriched of
expression within our annotations relative to unannotated regions.

The collection of ncRNA sequences is generally biased towards model
organisms \cite{Gardner:2010,Hoeppner:2012}. However, we have shown
that using data from well studied lineages such as mammals can also
result in quality annotations of sister taxa such as aves.

In summary, these results indicate we are in the very early phases of
determining the functions of many RNA families. This is illustrated by
the fact that the reported functions of some ncRNAs are
mammal-specific, yet are also found in bird genomes.

>>>>

\begin{abstract}
Here we present the results of a large-scale bioinformatic annotation
of non-coding RNA loci in 48 avian genomes. Our approach uses
probabilistic models of hand-curated families from the Rfam database
to infer conserved RNA families within each avian genome. We
supplement these annotations with predictions from the tRNA annotation
tool, tRNAscan-SE and microRNAs from miRBase.  We show that a number of
lncRNA-associated loci are conserved between birds and mammals,
including several intriguing cases where the reported mammalian lncRNA
function is not conserved in birds.  We also demonstrate extensive
conservation of classical ncRNAs (e.g., tRNAs) and more recently
discovered ncRNAs (e.g., snoRNAs and miRNAs) in birds. Furthermore, we
describe numerous ``losses'' of several RNA families, and attribute
these to genuine loss, divergence or missing data.  In particular, we
show that many of these losses are due to the challenges associated
with assembling Avian microchromosomes. These combined results illustrate the utility of
applying homology-based methods for annotating novel vertebrate
genomes.
\end{abstract}

@@

New sequencing technologies have dramatically expanded the rate at
which ncRNAs are discovered and their functions are determined
\cite{cech2014noncoding}. However, in order to determine the full
range of ncRNAs across multiple species we require multiple RNA
fractions (e.g. long and short), in multiple species, in multiple
developmental stages and tissues types. The costs of this approach are
still prohibitive in terms of researcher-time and
finances. Consequently, in this study we concentrate on bioinformatic
approaches, primarily we use homology-based methods (i.e. covariance models
(CMs)). However, we do validate the majority of these predictions
using RNA-seq. The bioinformatic approaches we use remain state of the art for ncRNA bioinformatic
analyses \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009} and have well
established sensitivity and specificity rates \cite{Freyhult:2007}.
For example, the CM based approach for annotating ncRNAs in genomes requires
reliable alignments and consensus secondary structures of
representative sequences of RNA families, many of which can be found
at Rfam
\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}. These
are used to train probabilistic models that score the likelihood that
a database sequence is generated by the same evolutionary processes as
the training sequences based upon both sequence and structural
information \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.  The
tRNAscan-SE software package uses CMs to accurately predict transfer
RNAs \cite{Lowe:1997,Chan:2009}.

Independent benchmarks of bioinformatic annotation tools have shown
that the CM approaches out-perform alternative methods
\cite{Freyhult:2007}, although their sensitivity can be limited for
rapidly evolving families such as vault RNAs or telomerase RNA
\cite{Menzel:09a}.

@@

The NBR2 lncRNA and BRCA1 gene share a bidirectional promotor
\cite{Xu:1997}. Both are expressed in a broad range of
tissues. Extensive research on BRCA1 has shown that it is involved in
DNA repair \cite{Moynahan:1999}. The function of NBR2 remains unknown,
yet its conservation across the vertebrates certainly implies a
function (See Figure~\ref{fig:1}). We note that the function for this
locus may be at the DNA level, however, function at the RNA level
cannot be ruled out at this stage.

@@

In order to determine to what extent the absence of some ncRNAs from
the infernal-based annotation is caused by sequence divergence beyond
the thresholds of the Rfam CMs, we complemented our analysis by
dedicated searches for a few of these RNA groups. Our ability to find
additional homologs for several RNA families that fill gaps in the
abundance matrices (Figure~1) strongly suggests that conspicuous
absences, in particular of LUCA and LECA RNAs, are caused by
incomplete data in the current assemblies and sequence divergence
rather then genuine losses.

@@

In this work we have provided a comprehensive annotation of non-coding
RNAs in genome sequences using homology-based methods. The
homology-based tools have distinct advantages over experimental-based
approaches as not all RNAs are expressed in any particular tissue-type
or developmental-stage, in fact some RNAs have extremely specific
expression profiles, e.g. the lsy-6 microRNA \cite{Johnston:2003}.  We
have identified previously unrecognised conservation of ncRNAs in
avian genomes and some surprising ``losses'' of otherwise well
conserved ncRNAs. We have shown that most of these losses are due to
difficulties assembling avian microchromosomes rather than \emph{bona
  fide} gene loss. A large fraction of our annotations have been
confirmed using RNA-seq data, which also showed a 7-fold enrichment of
expression within our annotations relative to unannotated regions.

The collection of ncRNA sequences is generally biased towards model
organisms \cite{Gardner:2010,Hoeppner:2012}. However, we have shown
that using data from well studied lineages such as mammals can also
result in quality annotations of sister taxa such as Aves.

In summary, these results indicate we are in the very early phases of
determining the functions of many RNA families. This is illustrated by
the fact that the reported functions of some ncRNAs are
mammal-specific, yet these are also found in bird genomes.
